Reassessing the role of theory and machine learning in forecasting civil conflict

### 2020-09-18 paper.Rmd
back to you, andy

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves justified as easing interpretation. Smoothing is used to produce all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), for a total of 78 items, in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.} (Both of the ambiguous cases include Blair as coauthor.) The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.

INSERT SMALL STUDY HERE WITH DETAILS.

@@

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.

@@

There is another type of theory we can define ostensibly. It consists of taking a variety of variables that putatively relate to another (dependent) variable and running a non-parametric model such as a logistic regression with this set of variables.  This even weaker form of theory is the most widespread use of theory in the contemporary social sciences, and it is the type of theory that is employed in B\&S:

It is not a procedure.  They actually create a model with ten right-hand side variables that are supposed to capture a complicated repression-dissent dynamic. There is wide-ranging literature on this dynamic that could justify many specifications.  As such, their baseline comparison is an unfortunate standard bearer for strong theory.

>>>>

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves justified as easing interpretation. Smoothing is used to produce all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), for a total of 78 items, in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.} (Both of the ambiguous cases include Blair as coauthor.) The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.

@@

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.

@@

Most social science research uses a weak form of theory.  In the case of B\&S they base their theoretical thinking on research that argues that conflict is not the product of structural factors, but rather comes from an interaction of state repression and the changes in costs and benefits to citizens for engaging in anti-government conflict. Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars.  Operationally, this boils down to four types of concepts: demands, accommodations,
nonviolent repression, and low-level violence.
\begin{itemize}
\item Demands include events in which either opposition or rebels is the source and
government is the target and typically involve appeals for political or institutional
reform, policy change, or extension of rights.
\item Nonviolent repression and accommodation include events in which government
is the source and either opposition or rebels is the target. Nonviolent
repression includes threats, sanctions, restrictions on freedom of movement, or
rejections of group demands.
\item Low-level violence includes all events in which
(1) government is the source and opposition is the target, (2) opposition is the
source and government is the target, (3) government is the source and rebels are
the target, or (4) rebels are the source and government is the target.
\end{itemize}
These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article.

Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story. A strong theory gives more guidance on compelling choices than does a weak theory.  Accordingly, while the choice is plausible is not necessarily compelling.  Another choice might be equally plausible.  In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.    In practical terms there is a wide array of choices that are available to instantiate any particular theory.  A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated temporal dynamics.  One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.

MORE HERE ON GENERAL IDEAS.  ANDY ANY BULLET pOINTs TO ADD?
